---
title: "time series patterns/graphics"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(feasts)
library(tsibble)
library(readr)
library(tidyverse)
library(lubridate)
library(tsbox)
library(dygraphs)
```


```{r arrival_tsb, include=FALSE}

arrival <- read_csv(
  file = "data/ae_uk.csv") |> mutate(arrival_time= dmy_hm(arrival_time))

arrival %>% duplicated() %>% sum()#check duplicates

arrival_wd <- arrival %>% dplyr::distinct()# remove duplicates and get a distinct tibble

arrival_tsb <- arrival_wd %>% 
  as_tsibble(key = c(gender,type_injury), index = arrival_time, regular=FALSE)


arrival_hourly <- arrival_tsb %>% group_by(gender, type_injury)  %>% 
  index_by(time = lubridate::floor_date(arrival_time, "1 hour")) %>%
  summarise(arrival= n(), .groups = "drop") %>% fill_gaps(arrival=0)

total_arrival_hourly <- arrival_tsb %>%  group_by(gender) %>% 
  index_by(time = lubridate::floor_date(arrival_time, "1 hour")) %>%
  summarise(arrival= n(), .groups = "drop") %>% fill_gaps(arrival=0)


arrival_daily <- arrival_hourly %>%  group_by(gender, type_injury)  %>%
  index_by(date=lubridate::as_date(time)) %>% 
  summarise(arrival=sum(arrival))

total_arrival_daily <- arrival_hourly %>% 
  index_by(date=lubridate::as_date(time)) %>% 
  summarise(arrival=sum(arrival))
```

# Lab 2: time series patterns/graphics

## time plot

##You can create time plot using `autoplot()` function

```{r ts-plot}
arrival_daily %>% feasts::autoplot() # create a time plot of daily data
arrival_hourly %>% autoplot() # create a time plot of hourly data

total_arrival_hourly %>% feasts::autoplot() 
total_arrival_daily %>% autoplot()
```

You can use filter_index or head() and tail() to select a subset of time series , try ?filter_index or google it


```{r filter-index}
total_arrival_daily %>% tsibble::filter_index("2016-02") %>% autoplot(arrival) # plot only February 2016 
total_arrival_daily %>% tsibble::filter_index("2016-02" ~ .) %>% autoplot(arrival) # plot from February 2016 till the end
total_arrival_daily %>% tsibble::filter_index(. ~ "2016-02") %>% autoplot(arrival) # plot from February 2016 till the end
total_arrival_hourly %>% tsibble::filter_index("2015-01-01") %>% autoplot(arrival) # plot only 2015-01-01
total_arrival_daily %>% head(n=100) %>% autoplot() # first 100 observations
total_arrival_daily %>% tail(n=100) %>% autoplot()  # last 100 observations

```

It could be really hard to visualise hourly or daily data, one possibility is to view a portion of the data, you can also use  interactive plots:

```{r admission-dygraph}
tsbox::ts_xts(total_arrival_daily) %>% 
  dygraph() %>% 
  dyRangeSelector(dateWindow = c("2010-01-01", "2010-02-01"))

tsbox::ts_xts(total_arrival_hourly) %>% 
  dygraph() %>% 
  dyRangeSelector(dateWindow = c("2010-01-01", "2010-01-02"))
```

You can also plot monthly, weekly, quarterly time series using ggplot2 package, see below for an example

```{r monthly-plot}

monthly_admissions <- total_arrival_daily %>% 
  index_by(month = yearmonth(date)) %>%
  summarise(arrival = sum(arrival))

  
monthly_admissions %>% 
  autoplot(arrival) +
  labs(y = "arrival", x="Month",
       title = "Monthly A&E arrival",
       subtitle = "UK hospital")
```

## Seasonal plots

use seasonal and subseries plots to check whether series contain seasonality 

```{r seasonal-plot}
total_arrival_daily %>% feasts::gg_season(arrival) 
total_arrival_daily %>% feasts::gg_subseries(arrival)
```

You can also try this with hourly series, you can change the "period= ..." argument to see seasonality

```{r multiple-seasonality}
total_arrival_hourly %>% feasts::gg_season(arrival, period = "day") 
total_arrival_daily %>% feasts::gg_season(arrival, period = "week")
```

```{r seasonal-period, options}
total_arrival_hourly %>% gg_season(arrival,period = "week")# change period 
```

Is there any seasonality in the daily time series? what about hourly and monthly?


How do you create a seasonal plot for the weekly, monthly and quarterly series series

```{r weekly-seasonality}
weekly_admissions <- total_arrival_daily %>% 
  index_by(week = yearweek(date)) %>%
  summarise(arrival = sum(arrival))
gg_season(weekly_admissions)
gg_subseries(weekly_admissions)
```

```{r monthly-seasonality}
gg_season(monthly_admissions)
gg_subseries(monthly_admissions)

```


```{r quarterly-seasonality}
quarterly_admissions <- total_arrival_daily %>% 
  index_by(quarter = yearquarter(date)) %>%
  summarise(arrival = sum(arrival))
gg_season(quarterly_admissions)
gg_subseries(quarterly_admissions)
```

## autocorrelation plot

```{r acf-calcul}
total_arrival_daily %>% feasts::gg_lag(arrival, lags = c(1:14), geom = "point")# create lag plots for 14 lags, from 1 to 14
total_arrival_daily %>% feasts::ACF(lag_max = 14)# compute autocorrelation function for 14 lags
```

plot the autocorrelation 

```{r acf-plot}
total_arrival_hourly %>% ACF(lag_max = 48) %>% autoplot()# plot acf
total_arrival_daily %>% ACF(lag_max = 14) %>% autoplot()# plot acf
```

You can have a time plot, acf and season plot in one single plot if you use `gg_tsdisplay()` function

```{r multiple-plot-, options}
# total_arrival_hourly %>% filter(gender == "female") %>%  gg_tsdisplay()#  check ?gg_tsdisplay
# total_arrival_daily %>% filter(gender == "female") %>% gg_tsdisplay()#  check ?gg_tsdisplay

total_arrival_hourly %>% filter(gender == "female") %>%  gg_tsdisplay()#  check ?gg_tsdisplay
total_arrival_daily  %>% gg_tsdisplay()#  check ?gg_tsdisplay


```

You can use ljung box to test whether ACF is significant, if p-value is small, << 0.05 then there is a significant autocorrelation:

```{r ljung-box-test}
total_arrival_daily %>% features(arrival, ljung_box)
```

What autocorrelation will tell us? Which key features could be highlighted by ACF?

## Make any graph using ggplot2 (optional)

You can create any graph that helps you to better understand data! I recommend you to look at the distributions of your variables, try geom_boxplot() , geom_histogram() and geom_density() which are helpful to better understand the variations

Here I tried to see if arrival of males or females is different over the weekend comparing to the weekday

```{r weekend}
weekend_an_weekday <- total_arrival_hourly %>% group_by(gender) %>% 
  summarise(arrival=sum(arrival)) %>% 
  mutate(
    Date=lubridate::as_date(time),
    hour=lubridate::hour(time),
    Day = lubridate::wday(time, label = TRUE),
    Weekend = (Day %in% c("Sun", "Sat"))) %>% 
  filter(gender =="female") 
weekend_an_weekday %>% ggplot(aes(x = hour, y = arrival)) +
  geom_line(aes(group=Date)) +
  facet_grid(Weekend ~., scales="free_y")
```

  
  